---
title: "CS_JG_MBIO612_Final_Proj"
output: pdf_document
date: "2022-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#add libraries
library(gdata)
library(reshape)
library(tidyverse)
library(ggplot2)
#add data
raw_data <- read.csv("https://raw.githubusercontent.com/CallieStephenson/CS-MBIO-612/main/Final_Proj_CS_JG/HI_ONLY_NOAA_PACIFIC_RAMP_FISH_SPC_2010_2017_SCI_DATA_MBIO612.csv")
```
## Chapter 1: Dataset Proposal
#Part 1: Tidy Data

At least 5 commands to tidy data
Tidy data:
Each variable must have its own column.
Each observation must have its own row.
Each value must have its own cell.

```{r Tidy Data}
#now we only want data from instantaneous (I) and noninstantaneous (N) observations
wd<-subset(raw_data, (raw_data$OBS_TYPE =="I"|raw_data$OBS_TYPE =="N"))
#and we only want to look at surveys from habitats with less than 50% sand
wd<-subset(wd, wd$SAND < 50.01)
#if we want to look at areas with high coral cover only
high_cc <- subset(wd, wd$HARD_CORAL > 49.99)
#lastly, we want to focus on surveys that took place in less than 18 m
wd<-subset(wd, wd$DEPTH_M < 18.01)

```
## Chapter 1: Dataset Proposal
## Part 2: Clean Data

At least five R commands to clean your data
Clean data:
No missing values
No impossible values

```{r pressure, echo=FALSE}
########Our chief questions with this dataset were: 
#How does coral cover impact species richness or biodiversity?
#How does depth impact species richness or biodiversity?
#Is there overall more species in the NWHI than in the MHI
#Are there more sharks in the NWHI than in the MHI
#*have not done yet*Are there more herbivores in the NWHI than in the MHI
###########
#To answer these questions, we created 3 Data Frames
#The first we made using lapply
#To show diversity in coding, we made a second using a for loop to display overall biomass for each survey
#A lot of the columns were not useful for us here, we made one data frame for Species Richness (using lapply) and another for Overall Biomass (using a for loop)
#Here is Species Richness using lapply:
Richness_by_Depth_and_Year <- data.frame( Region=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){raw_data$REGION[which(raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x])[1]]})), Island=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){raw_data$ISLAND[which(raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x])[1]]})), Depth=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){raw_data$MAX_DEPTH_M[which(raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x])[1]]})), Year=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){raw_data$OBS_YEAR[which(raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x])[1]]})), Region=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){raw_data$REGION[which(raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x])[1]]})), Species_Richness=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){length(unique(raw_data$SPECIES[raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x]]))})))

#######
#And here is a for loop to create Biomass for each survey:
raw_data$BiomassPerFish=raw_data$LW_A*((raw_data$SIZE_TL_CM*raw_data$LENGTH_CONVERSION_FACTOR)^raw_data$LW_B)
raw_data$BiomassPerObs=raw_data$BiomassPerFish*raw_data$COUNT

Biomass_for_Survey=data.frame(Site=NA,Total_Biomass=NA,Island=NA,Region=NA,Year=NA)

for(i in 1:length(unique(raw_data$SITE))){
  Biomass_for_Survey=rbind(Biomass_for_Survey,c(unique(raw_data$SITE)[i],sum(raw_data$BiomassPerObs[which(raw_data$SITE==unique(raw_data$SITE)[i])]),raw_data$ISLAND[which(raw_data$SITE==unique(raw_data$SITE)[i])][1],raw_data$REGION[which(raw_data$SITE==unique(raw_data$SITE)[i])][1],raw_data$OBS_YEAR[which(raw_data$SITE==unique(raw_data$SITE)[i])][1]))
}
Biomass_for_Survey=Biomass_for_Survey[-1,]
Biomass_for_Survey$Total_Biomass <- as.numeric(Biomass_for_Survey$Total_Biomass)

#Then this for loop makes an average chart that can be used in a barchart
##I BROKE THIS CODE DO NOT LOOK AT IT
AvgBiomassIslandYear=data.frame(Island=NA, Region=NA, Biomass=NA)
rm(i)
for(i in 1:length(unique(Biomass_for_Survey$Island))){
  AvgBiomassIslandYear=rbind(AvgBiomassIslandYear,c(unique(Biomass_for_Survey$Island)[i], (Biomass_for_Survey$Region)[i],mean(as.numeric(Biomass_for_Survey$Total_Biomass[Biomass_for_Survey$Island==unique(Biomass_for_Survey$Island)[i]]))))
}

AvgBiomassIslandYear=AvgBiomassIslandYear[-1,]
AvgBiomassIslandYear$Biomass=as.numeric(AvgBiomassIslandYear$Biomass)


#####
#Lastly, looking just at shark observations
Sharks=c("Carcharhinus amblyrhynchos","Carcharhinus galapagensis","Carcharhinus melanopterus","Triaenodon obesus","Sphyrna lewini")

SharkDF=raw_data[which(raw_data$TAXONNAME %in% Sharks),]
```
## Part 3: Data Visualizations
At least five ggplot2 data visualizations that describe your data. You can choose any data visualization you deem valuable to understand your data.

```{data vis}
#This shows by depth how much hard coral was at each site in the subset of areas with coral cover >50%, which makes it seem like depth was not a significant factor in coral cover
ggplot(data = high_cc) + 
     geom_point(mapping = aes(x = DEPTH_M, y = HARD_CORAL, color = REGION))

####################
# A quick boxblot of species richness by island
Richness_by_Island <- ggplot(Richness_by_Island_and_Year,aes(x = Island, y = Species_Richness)) + geom_boxplot ()

# A quick point of species richness by island
Richness_by_Depth <- ggplot(Richness_by_Depth_and_Year,aes(x=Depth, y =Species_Richness, color = Region)) + geom_point ()

# A quick bar chart of biomass by island
Biomass_by_Island <- ggplot(AvgBiomassIslandYear,aes(x = island, y = biomass)) + geom_col ()+ theme(axis.text.x=element_text(angle=90,hjust=1))

# A quick boxblot of biomass by island
Biomass_by_Island <- ggplot(Biomass_for_Survey,aes(x = Island, y = Total_Biomass)) + geom_boxplot ()
##SHOWS INSANE OUTLIERS

##################
## Sharks in MHI vs NWHI
Sharks <- ggplot(SharkDF, aes(x=ISLAND, y=COUNT, fill = REGION))+geom_col() + theme(axis.text.x=element_text(angle=90,hjust=1))+ facet_grid(. ~ REGION)
# LOOK AT HOW MANY MORE SHARKS THERE ARE IN THE NWHI
##################

```
## Chapter 2: Final Project
#Part 1: Data Wrangling

Common issues encountered when dealing with this type of data. Consider missing values; can you impute or should you drop them? Data may not be in the correct format, such as site names that need to be standardized or data that needs to be combined from two different files
```{r cars}
summary(cars)
```
## Chapter 2: Final Project
#Part 2: Exploratory data analysis

Useful visualizations for gaining a quick overview of the data. Such visualizations are typically perfect for detecting any residual anomalies in the data, or for simply observing any trends or differences between groups or samples
```{r cars}
summary(cars)
```
## Chapter 2: Final Project
#Part 3: Analysis and/or modeling

What types of analyses are common in this domain? Asking someone who specializes in this type of analysis is the easiest way to answer this question
```{r cars}
summary(cars)
```
## Chapter 2: Final Project
#Part 4: Insight communication

Using appropriate data visualization to summarize your results. Once again, I suggest consulting a colleague who is familiar with the types of visualization used with the data you are using.
```{r cars}
summary(cars)
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
