---
title: "CS_JG_MBIO612_Final_Proj"
output: pdf_document
date: "2022-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#CALLIE'S NOTES FOR JESS: I think we're in a pretty good place! I can write some code to do some kind of statistical analysis on the data - like 'does region make a significant difference in total biomass/biomass of sharks'. 

#If we want to do more cleaning/tidying/wrangling, I found this in Julie's code that might be handy?
#JZ: Manually replace the Island name for Pearl & Hermes and French Frigate shoals so they are called the universal "PHR" and "FFS" across datasheets
#wd<- within(wd, ISLAND[ISLAND == "Pearl & Hermes"] <- "PHR")
#wd<- within(wd, ISLAND[ISLAND == "French Frigate"] <- "FFS") 

#add libraries
library(gdata)
library(reshape)
library(tidyverse)
library(ggplot2)
#add data
raw_data_1 <- read.csv("https://raw.githubusercontent.com/CallieStephenson/CS-MBIO-612/main/Final_Proj_CS_JG/HI_ONLY_NOAA_PACIFIC_RAMP_FISH_SPC_2010_2017_SCI_DATA_MBIO612.csv")
#We got this data from a member in our lab who stressed ot us the importance of not using observation types other than Instantaneous (I) and Noninstantaneous (N). Though this dataset should only be SPC (standardized point count) data, some tow surveys were included that could bias biomass estimates.
raw_data <- subset(raw_data_1, (raw_data_1$OBS_TYPE =="I"|raw_data_1$OBS_TYPE =="N"))
```
## Chapter 1: Dataset Proposal
#Part 1: Tidy Data

At least 5 commands to tidy data
Tidy data:
Each variable must have its own column.
Each observation must have its own row.
Each value must have its own cell.

NOAA dataset was already pretty tidy, some code is written to untidy the data, and the code following is written to re-tidy the data. This dataset was also very large, so we chose to tidy it by creating new dataframes that included only the information that we needed to run specific analyses and to remove excess clutter. 

```{r Tidy Data}
#subset the large dataframe to include only columns we will use for analysis
new_data <- raw_data[ , ! names(raw_data) %in% c("DATAROWID", "REGION", "SITE", "DIVER", "REPLICATEID", "REP", "HABITAT_CODE", "CURRENT_STRENGTH", "VISIBILITY_M", "MIN_DEPTH_M", "COMPLEXITY", "SUBSTRATE_HEIGHT_0", "SUBSTRATE_HEIGHT_20", "SUBSTRATE_HEIGHT_50", "SUBSTRATE_HEIGHT_100", "SUBSTRATE_HEIGHT_150", "MAX_HEIGHT", "URCHIN_DACOR", "BORING_URCHIN_DACOR", "COMMON_FAMILY", "FAMILY", "CONSUMER_GROUP", "OBS_TYPE")]

#split dataframe so that each row becomes 2 rows, REEF_ZONE and DEPTH_BIN become variables rather than columns
data_long <- pivot_longer(data = new_data,
                          cols = c("REEF_ZONE","DEPTH_BIN"))

#place 2 variables into columns side-by-side again to reduce length of dataframe
data_wide <- pivot_wider(data=data_long)

#reduce the number of sig figs on Latitude to match number of sig figs on Longitude
new_data[,'LATITUDE']=round(new_data['LATITUDE'],4)


#now we only want data from instantaneous (I) and noninstantaneous (N) observations
wd<-subset(raw_data, (raw_data$OBS_TYPE =="I"|raw_data$OBS_TYPE =="N"))

#and we only want to look at surveys from habitats with less than 50% sand
wd<-subset(wd, wd$SAND < 50.01)

#if we want to look at areas with high coral cover only
high_cc <- subset(wd, wd$HARD_CORAL > 49.99)

#lastly, we want to focus on surveys that took place in less than 18 m
wd<-subset(wd, wd$DEPTH_M < 18.01)

```
## Chapter 1: Dataset Proposal
## Part 2: Clean Data

At least five R commands to clean your data
Clean data:
No missing values
No impossible values

To clean our data, we created two data frames based off the information we wanted to look at from the larger dataset. The code below makes one dataframe for Biomass found at each survey, and species richness of each survey. We used lapply and a for loop to show a variety of knowledge. We also subsetted this data for shark-specific data analysis.

As some values were missing Depth information, these needed to be excluded from analysis, and we cleaned the data for this.

There are extreme outliers in the biomass data, where there are large observations on dives (20 jacks of 120 cm at French Frigate Shoals) that wildly skew some of the data, and these are removed in some analyses as is relevant. It is important to note that this data had already been extensively QA/QC'd and we do not expect these to be incorrect daapoints, but rather natural processes that are not well captured in the survey design and present as anomalies.

```{r pressure, echo=FALSE}

###########
#For the first section of this projet we will focus on species richness
#A lot of the columns were not useful for us here, we made one data frame for Species Richness (using lapply)
Richness_by_Depth_and_Year <- data.frame( Region=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){raw_data$REGION[which(raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x])[1]]})), Island=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){raw_data$ISLAND[which(raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x])[1]]})), Depth=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){raw_data$MAX_DEPTH_M[which(raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x])[1]]})), Year=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){raw_data$OBS_YEAR[which(raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x])[1]]})), Region=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){raw_data$REGION[which(raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x])[1]]})), Species_Richness=unlist(lapply(c(1:length(unique(raw_data$SITEVISITID))),function(x){length(unique(raw_data$SPECIES[raw_data$SITEVISITID==unique(raw_data$SITEVISITID)[x]]))})))

#######
```
## Chapter 1: Dataset Proposal
## Part 3: Data Visualizations
At least five ggplot2 data visualizations that describe your data. You can choose any data visualization you deem valuable to understand your data.

```{data vis}
#This shows by depth how much hard coral was at each site in the subset of areas with coral cover >50%, which makes it seem like depth was not a significant factor in coral cover. This seeks to validate the survey design - there should be no correlation between the two variables.
ggplot(data = high_cc) + 
     geom_point(mapping = aes(x = DEPTH_M, y = HARD_CORAL, color = REGION))

####################
# A quick boxblot of species richness by island
Richness_by_Island <- ggplot(Richness_by_Island_and_Year,aes(x = Island, y = Species_Richness)) + geom_boxplot ()

####################
# A quick boxblot of species richness by region. We believe from Toonen et al 2010 that the NWHI are reproductively separate from the MHI, and wanted to see the overall difference in richness between the two.
Richness_by_Island <- ggplot(Richness_by_Island_and_Year,aes(x = Region, y = Species_Richness)) + geom_boxplot ()

##################
# A quick point of species richness by depth
Richness_by_Depth <- ggplot(Richness_by_Depth_and_Year,aes(x=Depth, y =Species_Richness, color = Region)) + geom_point ()

###################
#A Scatterplot of richness over time to look for any trends. Note that the NWHI and MHI experienced a bleaching event from 2014-2015, so we wanted to see if this might have a broad impact on species richenss 
Richness_over_time <- ggplot(Richness_by_Depth_and_Year, aes(x=Year, y =Species_Richness)) + geom_point()

```
## Chapter 2: Final Project
#Part 1: Data Wrangling

Common issues encountered when dealing with this type of data. Consider missing values; can you impute or should you drop them? Data may not be in the correct format, such as site names that need to be standardized or data that needs to be combined from two different files
```{Biomass}
#add libraries
library(gdata)
library(reshape)
library(tidyverse)
library(ggplot2)
#add data
raw_data <- read.csv("https://raw.githubusercontent.com/CallieStephenson/CS-MBIO-612/main/Final_Proj_CS_JG/HI_ONLY_NOAA_PACIFIC_RAMP_FISH_SPC_2010_2017_SCI_DATA_MBIO612.csv")
#We got this data from a member in our lab who stressed ot us the importance of not using observation types other than Instantaneous (I) and Noninstantaneous (N). Though this dataset should only be SPC (standardized point count) data, some tow surveys were included that could bias biomass estimates.
raw_data <- subset(raw_data_1, (raw_data_1$OBS_TYPE =="I"|raw_data_1$OBS_TYPE =="N"))

#Here is a for loop to create Biomass for each survey:
raw_data$BiomassPerFish=raw_data$LW_A*((raw_data$SIZE_TL_CM*raw_data$LENGTH_CONVERSION_FACTOR)^raw_data$LW_B)
raw_data$BiomassPerObs=raw_data$BiomassPerFish*raw_data$COUNT
#
Biomass_for_Survey=data.frame(Site=NA,Total_Biomass=NA,Island=NA,Region=NA,Year=NA,Depth=NA)

for(i in 1:length(unique(raw_data$SITE))){
  Biomass_for_Survey=rbind(Biomass_for_Survey,c(unique(raw_data$SITE)[i], sum(raw_data$BiomassPerObs[which(raw_data$SITE==unique(raw_data$SITE)[i])]), raw_data$ISLAND[which(raw_data$SITE==unique(raw_data$SITE)[i])][1], raw_data$REGION[which(raw_data$SITE==unique(raw_data$SITE)[i])][1], raw_data$OBS_YEAR[which(raw_data$SITE==unique(raw_data$SITE)[i])][1], raw_data$DEPTH_M[which(raw_data$SITE==unique(raw_data$SITE)[i])][1]))
}
Biomass_for_Survey=Biomass_for_Survey[-1,]
Biomass_for_Survey$Total_Biomass <- as.numeric(Biomass_for_Survey$Total_Biomass)
#####
#Pulling out data for shark-specific analyses
Sharks=c("Carcharhinus amblyrhynchos","Carcharhinus galapagensis","Carcharhinus melanopterus","Triaenodon obesus","Sphyrna lewini")

SharkDF=raw_data[which(raw_data$TAXONNAME %in% Sharks),]

#####
#Looking at shark biomass over time
Shark_Biomass_per_Survey=data.frame(Site=NA,Total_Biomass=NA,Island=NA,Region=NA,Year=NA,Depth=NA,Richness=NA)

for(i in 1:length(unique(SharkDF$SITE))){
  Shark_Biomass_per_Survey=rbind(Shark_Biomass_per_Survey,c(unique(SharkDF$SITE)[i], sum(SharkDF$BiomassPerObs[which(SharkDF$SITE==unique(SharkDF$SITE)[i])]), SharkDF$ISLAND[which(SharkDF$SITE==unique(SharkDF$SITE)[i])][1], SharkDF$REGION[which(SharkDF$SITE==unique(SharkDF$SITE)[i])][1], SharkDF$OBS_YEAR[which(SharkDF$SITE==unique(SharkDF$SITE)[i])][1], SharkDF$DEPTH_M[which(SharkDF$SITE==unique(SharkDF$SITE)[i])][1], length(unique((SharkDF$SPECIES[which(SharkDF$SITE==unique(SharkDF$SITE)[i])])))))
}
Shark_Biomass_per_Survey=Shark_Biomass_per_Survey[-1,]
Shark_Biomass_per_Survey$Total_Biomass <- as.numeric(Shark_Biomass_per_Survey$Total_Biomass)
```
<<<<<<< HEAD
=======
## Part 3: Data Visualizations
At least five ggplot2 data visualizations that describe your data. You can choose any data visualization you deem valuable to understand your data.

```{data vis}
#This shows by depth how much hard coral was at each site in the subset of areas with coral cover >50%, which makes it seem like depth was not a significant factor in coral cover, which shows the sampling design is robust.
ggplot(data = high_cc) + 
     geom_point(mapping = aes(x = DEPTH_M, y = HARD_CORAL, color = REGION))

####################
# A quick boxblot of species richness by island
Richness_by_Island <- ggplot(Richness_by_Island_and_Year,aes(x = Island, y = Species_Richness)) + geom_boxplot ()

# A quick point of species richness by island
Richness_by_Depth <- ggplot(Richness_by_Depth_and_Year,aes(x=Depth, y =Species_Richness, color = Region)) + geom_point ()

# A quick bar chart of biomass by island
Biomass_by_Island <- ggplot(Biomass_for_Survey,aes(x = Island, y = Total_Biomass)) + geom_boxplot () + theme(axis.text.x=element_text(angle=90,hjust=1))+ facet_grid(. ~ Region, scales = "free")
Biomass_by_Island

# A quick boxblot of biomass by island
Biomass_by_Island <- ggplot(Biomass_for_Survey,aes(x = Island, y = Total_Biomass)) + geom_boxplot ()+ theme(axis.text.x=element_text(angle=90,hjust=1))
##SHOWS INSANE OUTLIERS

##################
## Sharks in MHI vs NWHI
Sharks <- ggplot(SharkDF, aes(x=ISLAND, y=COUNT, fill = REGION))+geom_col() + theme(axis.text.x=element_text(angle=90,hjust=1))+ facet_grid(. ~ REGION)
# LOOK AT HOW MANY MORE SHARKS THERE ARE IN THE NWHI

##################

```
## Chapter 2: Final Project
#Part 1: Data Wrangling

Common issues encountered when dealing with this type of data. Consider missing values; can you impute or should you drop them? Data may not be in the correct format, such as site names that need to be standardized or data that needs to be combined from two different files
```{r cars}
### The dataset we used had some NA's in the depth column, but with 1898 surveys, there is still strong ability to infer when removing these columns. If depth was going to be used in one of our analyses, we could remove the rows from the dataset that had NA's for depth by using the following code: 

#drop NAs from depth column, this will remove entire row if depth is an NA, save as new dataframe
data_depth <- new_data %>%
               drop_na(MAX_DEPTH_M)

### Other issues that may arise using this type of data is having two variables in the same column. The NOAA dataset had both genus and species in the same column. If we wanted those in separate columns we would use the code below:
gen_sp <- new_data %>% separate(TAXONNAME, c('GENUS', 'SP'))



```
## Chapter 2: Final Project
#Part 2: Exploratory data analysis

Useful visualizations for gaining a quick overview of the data. Such visualizations are typically perfect for detecting any residual anomalies in the data, or for simply observing any trends or differences between groups or samples
<<<<<<< HEAD
```{r cars}
#These visualizations just make sure that no outliers are in the data

# A quick boxblot of biomass by island
Biomass_by_Island <- ggplot(Biomass_for_Survey,aes(x = Island, y = Total_Biomass)) + geom_boxplot () + theme(axis.text.x=element_text(angle=90,hjust=1))+ facet_grid(. ~ Region, scales = "free")
Biomass_by_Island

##SHOWS INSANE OUTLIERS

##################
## Sharks in MHI vs NWHI
Sharks <- ggplot(SharkDF, aes(x=ISLAND, y=COUNT, fill = REGION))+geom_col() + theme(axis.text.x=element_text(angle=90,hjust=1))+ facet_grid(. ~ REGION, scales="free")
Sharks
# LOOK AT HOW MANY MORE SHARKS THERE ARE IN THE NWHI

### First, we would want to take a subset of the larger dataframe to create one with just the variables we are exploring. In this case, we are interested in shark density by reef zone. 

#create new dataframe that includes just the shark species and the reef zone they are found in
reef_zone <- data.frame(data1$REEF_ZONE, data1$SPECIES, data1$TAXONNAME, data1$COUNT)
reef_zone_sharks <- reef_zone[reef_zone$data1.TAXONNAME %in% c('Carcharhinus amblyrhynchos', 'Carcharhinus galapagensis', 'Carcharhinus melanopterus','Triaenodon obesus','Sphyrna lewini'),] 

#create a bargraph to see the distribution of shark species by reef zone
ggplot(data = reef_zone_sharks, aes(fill=data1.TAXONNAME, y=data1.COUNT, x=data1.REEF_ZONE)) +
  geom_bar(position='dodge', stat='identity') +
  ggtitle('Shark Species Frequency By Reef Location') +
  xlab('Reef Zone') +
  ylab('Frequency')

###Additionally, we could create a scatterplot to visualize the trend in total length of a specific shark by depth

depth_df <- data.frame(data_depth$MAX_DEPTH_M, data_depth$TAXONNAME, data_depth$SIZE_TL_CM)

depth_gray_sharks <- depth_df[depth_df$data_depth.TAXONNAME %in% c('Carcharhinus amblyrhynchos'),] 

#plot the data in a scatterplot, the loess line gives us a gray confidence interval band around it
ggplot(depth_gray_sharks, aes(x=data_depth.MAX_DEPTH_M, y=data_depth.SIZE_TL_CM)) + geom_point() + geom_smooth(method=loess) +
  ggtitle('Relationship of Total Length and Depth of Carcharhinus amblyrhynchos') +
  xlab ('Observation Depth (M)') +
  ylab ('Shark Total Length (CM)')

##We can see that there is no relationship between depth and total length of Carcharhinus amblyrhynchos. Most points are clustered around the straight trendline and the confidence interval band is narrow, indicating there are likely no outlier values.

###If we create the same plot, but for Sphyrna lewini, we can see that there is an outlier value that pulls the middle of the trendline up so that it is curved. Additionally, the confidence interval band is wide.

depth_hammerhead <- depth_df[depth_df$data_depth.TAXONNAME %in% c('Sphyrna lewini'),] 

ggplot(depth_hammerhead, aes(x=data_depth.MAX_DEPTH_M, y=data_depth.SIZE_TL_CM)) + geom_point() + geom_smooth(method=loess) +
  ggtitle('Relationship of Total Length and Depth of Sphyrna lewini') +
  xlab ('Observation Depth (M)') +
  ylab ('Shark Total Length (CM)')

#Jess do you know what this is doing here: ">>>>>>> 1334c1b41edb8d81a1f82c1b242f62ae8ffc10f4"
```
## Chapter 2: Final Project
#Part 3: Analysis and/or modeling

What types of analyses are common in this domain? Asking someone who specializes in this type of analysis is the easiest way to answer this question
```{Average Biomass for Each Year}
#Our earlier histograms showed some CRAZY outliers in the Biomass data. Here, 'outliers' is the biomass outlier values and 'outliers_in_data' tells us which row they come at in the dataframe
outliers <- boxplot.stats(Biomass_for_Survey$Total_Biomass)$out
outliers_in_data <- which(Biomass_for_Survey$Total_Biomass %in% c(outliers))

#since there are 182 outliers in the 1898 surveys, that's about 1% of the surveys (is this just how outliers works??). We can see the outlier data here:
head(Biomass_for_Survey[outliers_in_data,])

#Average biomasss for each island is something that is common to look at a quick snapshot of trends in biomass over time.
AvgBiomassIslandYear=data.frame(island=NA,biomass=NA)
rm(i)
for(i in 1:length(unique(Biomass_for_Survey$Island))){
  AvgBiomassIslandYear=rbind(AvgBiomassIslandYear,c(unique(Biomass_for_Survey$Island)[i],mean(as.numeric(Biomass_for_Survey$Total_Biomass[Biomass_for_Survey$Island==unique(Biomass_for_Survey$Island)[i]]))))
}
AvgBiomassIslandYear=AvgBiomassIslandYear[-1,]
AvgBiomassIslandYear$biomass=as.numeric(AvgBiomassIslandYear$biomass)

#####
```
## Chapter 2: Final Project
#Part 4: Insight communication

Using appropriate data visualization to summarize your results. Once again, I suggest consulting a colleague who is familiar with the types of visualization used with the data you are using.
```{r cars}
# A quick bar chart of biomass by island
Biomass_by_Island <- ggplot(AvgBiomassIslandYear,aes(x = island, y = biomass)) + geom_col ()+ theme(axis.text.x=element_text(angle=90,hjust=1))
Biomass_by_Island
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
