---
title: "CS_JG_MBIO612_Final_Proj"
output: pdf_document
date: "2022-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#add libraries
library(gdata)
library(reshape)
library(tidyverse)
library(ggplot2)
#add data
raw_data <- read.csv("https://raw.githubusercontent.com/CallieStephenson/CS-MBIO-612/main/Final_Proj_CS_JG/HI_ONLY_NOAA_PACIFIC_RAMP_FISH_SPC_2010_2017_SCI_DATA_MBIO612.csv")
```
## Chapter 1: Dataset Proposal
#Part 1: Tidy Data

At least 5 commands to tidy data
Tidy data:
Each variable must have its own column.
Each observation must have its own row.
Each value must have its own cell.

```{r Tidy Data}
#now we only want data from instantaneous (I) and noninstantaneous (N) observations
wd<-subset(raw_data, (raw_data$OBS_TYPE =="I"|raw_data$OBS_TYPE =="N"))
#and we only want to look at surveys from habitats with less than 50% sand
wd<-subset(wd, wd$SAND < 50.01)
#if we want to look at areas with high coral cover only
high_cc <- subset(wd, wd$HARD_CORAL > 49.99)
#lastly, we want to focus on surveys that took place in less than 18 m
wd<-subset(wd, wd$DEPTH_M < 18.01)
```
## Chapter 1: Dataset Proposal
## Part 2: Clean Data

At least five R commands to clean your data
Clean data:
No missing values
No impossible values

```{r pressure, echo=FALSE}
plot(pressure)
```
## Part 3: Data Visualizations
At least five ggplot2 data visualizations that describe your data. You can choose any data visualization you deem valuable to understand your data.

```{r cars}
#This shows by depth how much hard coral was at each site in the subset of areas with coral cover >50%, which makes it seem like depth was not a significant factor in coral cover
ggplot(data = high_cc) + 
     geom_point(mapping = aes(x = DEPTH_M, y = HARD_CORAL, color = REGION))
```
## Chapter 2: Final Project
#Part 1: Data Wrangling

Common issues encountered when dealing with this type of data. Consider missing values; can you impute or should you drop them? Data may not be in the correct format, such as site names that need to be standardized or data that needs to be combined from two different files
```{r cars}
summary(cars)
```
## Chapter 2: Final Project
#Part 2: Exploratory data analysis

Useful visualizations for gaining a quick overview of the data. Such visualizations are typically perfect for detecting any residual anomalies in the data, or for simply observing any trends or differences between groups or samples
```{r cars}
summary(cars)
```
## Chapter 2: Final Project
#Part 3: Analysis and/or modeling

What types of analyses are common in this domain? Asking someone who specializes in this type of analysis is the easiest way to answer this question
```{r cars}
summary(cars)
```
## Chapter 2: Final Project
#Part 4: Insight communication

Using appropriate data visualization to summarize your results. Once again, I suggest consulting a colleague who is familiar with the types of visualization used with the data you are using.
```{r cars}
summary(cars)
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
