### Question 0

*  Draw a concepts diagram that uses all the following Github terminology 
   * Include any missing keywords that will simplify the concepts diagrams 


`Push`, `Repository`, `Clone`, `Pull`, `Pull Request`, `Branch`, `Merging`, `Github Client`, `README file`, `Private or Public`

```{r}
# Add your photo here
![Concept_Map_2](https://github.com/CallieStephenson/CS-MBIO-612/commit/3b3e50b929e45d4fccff00d476f9e05824a44eca)

```

#### Question 1

`matrix(c(1,2,3,4,5,6) , nrow = 3)`

* Running the expression produces the following matrix

|  |  |
|---|---|
| 1 | 4 |
| 2 | 5 |
| 3 | 6 |

* How can you modify the call to `matrix()` to produce the following matrix instead?

|   | col_1 |  col_2 |
| --- |---|---|
| row_1 | 1 | 2 |
| row_2 | 3 | 4 |
| row_3 | 5 | 6 |

* Note that you need to name the columns (col_1 and col_2) and name the rows (row_1, row_2, row_3)

Hint: Use the `?` symbol to invoke the matrix documentation

```{r}
q1_matrix <- matrix(c(1,2,3,4,5,6) , nrow = 3)
rownames(q1_matrix) = c("row_1", "row_2","row_3")
colnames(q1_matrix) = c("col_1", "col_2")
print(q1_matrix)
# Add your photo here
```

#### Question 1

`matrix(c(1,2,3,4,5,6) , nrow = 3)`

* Running the expression produces the following matrix

|  |  |
|---|---|
| 1 | 4 |
| 2 | 5 |
| 3 | 6 |

* How can you modify the call to `matrix()` to produce the following matrix instead?

|   | col_1 |  col_2 |
| --- |---|---|
| row_1 | 1 | 2 |
| row_2 | 3 | 4 |
| row_3 | 5 | 6 |

* Note that you need to name the columns (col_1 and col_2) and name the rows (row_1, row_2, row_3)

Hint: Use the `?` symbol to invoke the matrix documentation

```{r}
# Write your answer here
q1_matrix <- matrix(c(1,2,3,4,5,6) , nrow = 3)
rownames(q1_matrix) = c("row_1", "row_2","row_3")
colnames(q1_matrix) = c("col_1", "col_2")
print(q1_matrix)
```

#### Question 2

* Load then sort the airquality data frame on its `Temp` and `Solar.R` columns in reverse order (largest to smallest values)
  * The function to sort a data frame is called order
* Display only the first 15 lines of your table

```{r}
# Write your answer here
data("airquality")
decreasing_temp_decreasing_sol <- order(-airquality$Temp,-airquality$Solar.R)
head(airquality[decreasing_temp_decreasing_sol,],n=15)
```

#### Question 3

* Sort the airquality data frame on its `Temp` in decreasing order and `Solar.R` in increasing order
* Display only the first 15 lines of your table

```{r}
# Write your answer here
data("airquality")
decreasing_temp_increasing_sol <- order(-airquality$Temp,airquality$Solar.R)
head(airquality[decreasing_temp_increasing_sol,],n=15)
```

#### Question 4

* There are various ways to select a subset of observations from a data frame.

* Consult your [R Reference Card](https://cran.r-project.org/doc/contrib/Baggott-refcard-v2.pdf), see `Data Selection and Manipulation` section.
  * What operations can you use to select all observations where the temperature is 72. Give at least two different answers to this question


```{r}
# Write your answer here
data("airquality")
Temp_72_obvs <- airquality$Temp==72
airquality[Temp_72_obvs,]
#And
subset(airquality,Temp==72)
#A third
airquality[which(airquality$Temp==72),]

```

#### Question 6

* You may have noticed when working with the `airqulity` data that some values show as `NA`
 * `NA` stands for not available, or missing values.
* A major part of data wrangling consists of cleaning missing values by either:
  * Dropping the lines that have missing values
   * Sometimes we can drop the column with missing values if the column is made of predominantly missing values
  * Imputing the missing values, which uses educated guesses (or more complex algorithms) to fill the missing values
  
* Find and remove all rows that are missing values for the `Solar.R` or `Ozone` variables
* Save the cleaned data to a new data frame called airquality_no_na
  * How many lines have been removed?

```{r}
# Write your answer here
#Find and remove all rows that are missing values for the `Solar.R` or `Ozone` variables:
airquality_no_na <- airquality[(!is.na(airquality$Ozone))&(!is.na(airquality$Solar.R)),]

#How many lines have been removed:
nrow(airquality) - nrow(airquality_no_na)
#42
```

#### Question 7
* Let's use a different strategy and impute the missing value.
  * replace the missing values for Solar.R using that month's average.
  * Example:
    * The missing value for line 6 should be replaced with the average for month 5.
    * The missing value for line 97 should be replaced with the average for month 8.

```{r}
# Write your answer here
#find months without Solar.R
missing_Solar.R <- is.na(airquality$Solar.R)
airquality[missing_Solar.R,]

#fix may
solar.r_may <- subset(airquality, Month==5)
mean_may <- mean(solar_r_may, na.rm = TRUE)
solar.r_may$Solar.R[is.na(solar.r_may$Solar.R)]<-mean_may
solar.r_may

#fix august
solar.r_august <- subset(airquality, Month==8)
mean_august <- mean(solar_r_august, na.rm = TRUE)
solar.r_august$Solar.R[is.na(solar.r_august$Solar.R)]<-mean_august

#put it back together
june <- subset(airquality, Month == 6)
july <- subset(airquality, Month == 7)
sept <- subset(airquality, Month == 9)
q7_airquality <- rbind(solar.r_may, june, july, solar.r_august, sept)

#displayed for you
head(q7_airquality)


```

